Temporal Dynamics of Normalization Reweighting

### 2022-09-30 manuscript.Rmd
Minor changes to keep track of rejected trials.

# reasonably compact code to check which packages are installed, install the missing ones, and activate all
packagelist <- c('knitr','remotes','tictoc','R.matlab','bookdown','grImport','tiff','pals','ez','gtools') # list of CRAN packages
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
if (!'osfr' %in% installed.packages()[,1]){remotes::install_github("centerforopenscience/osfr")}
if (!'FourierStats' %in% installed.packages()[,1]){remotes::install_github("bakerdh/FourierStats")}
packagelist <- c(packagelist,'osfr','FourierStats')
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

osf_auth(token = '3dEYuhZNmwWbG3xuhRIiVRo0T2oniOkEP3Ip8i1LPG4PNjeTRln54eGNAG7oyTT9xozWwJ')

@@

    # compare each cluster to the null distribution, retain the significant ones
    clusterps <- 0*(1:nclusts)
    if (length(nulldist)==nresamples){
    for (cc in 1:nclusts){
      i <- which(abs(nulldist)>abs(sumtvals[cc]))
      if (!isempty(i)){clusterps[cc] <- length(i)/nresamples}
      if (clusterps[cc]<clustthresh){
        sigclustercount <- sigclustercount + 1
        clusterpoints[[sigclustercount]] <- clusterlist[[cc]]
        clusterpout[sigclustercount] <- clusterps[cc]
        }
    }}
  }
clustout <- NULL
clustout$clusterpoints <- clusterpoints
clustout$nclusters <- sigclustercount
clustout$pvals <- clusterpout

@@

# check if directories exist to store data files (ignored by git)
if (!dir.exists('temp/')){dir.create('temp/')}
if (!dir.exists('temp/figures/')){dir.create('temp/figures/')}
if (!dir.exists('temp/VilidaiteProcessed/')){dir.create('temp/VilidaiteProcessed/')}
if (!dir.exists('temp/raw/')){dir.create('temp/raw/')}
if (!dir.exists('temp/raw/temp/')){dir.create('temp/raw/temp/')}
if (!dir.exists('temp/MEGdata/')){dir.create('temp/MEGdata/')}
if (!dir.exists('temp/EEGprocessed/')){dir.create('temp/EEGprocessed/')}

@@

Normalization itself is not a single process. Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation [@Heeger1992; @Foley1994], eye-of-origin [@Legge1979; @Sengpiel1994] and spatial position [@Petrov2005; @Cannon1991]. At present there is evidence of normalization reweighting between stimuli with orthogonal orientations [@Aschner2018], and adjacent spatial positions [@Yiltiz2020]. We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways. This is plausible, given that suppression within and between the eyes has different spatiotemporal tuning [@Meese2009], and dichoptic masking can be reduced by adapting to the mask [@Sengpiel2005; @Li2005; @Baker2007], unlike within-eye masking. Dynamic fluctuations in interocular suppression are a key feature of binocular rivalry [@Wilson2003; @Alais2010], for which autism-related differences have also been reported [@Robertson2013].

@@

Our recent work has identified a potential autism-related difference in gain control using steady-state EEG methods. @Vilidaite2018 measured contrast response functions for flickering stimuli in a large group of neurotypical adults, who were split by their autism quotient [AQ, see @Baron-Cohen2001] score. The responses were comparable between high and low AQ groups at the flicker frequency of the stimulus. However at the second harmonic frequency (twice the flicker frequency), the high AQ group showed weaker responses than the low AQ group. This effect was replicated in an independent sample of 12 adults with an autism diagnosis (compared with neurotypical controls). Because second harmonic responses are caused by nonlinear interactions in the visual system (they are absent in a linear system), and divisive suppression is a nonlinearity, this potentially implicates differences in suppression in autism. Furthermore, there appears to be a developmental trajectory, as autistic children tested as part of the same study [@Vilidaite2018] showed weaker responses at both the first and second harmonic frequencies.

@@

@@

# datadir <- '/Volumes/Seagate/EEG data/UG2014/tar/'
# tempdir <- '/Volumes/Seagate/EEG data/UG2014/tar/temp/'
datadir <- 'temp/raw/'
tempdir <- 'temp/raw/temp/'

@@

      for (tr in 1:counter) {
        cond <- EEGdata$Trigger[triggertimes[tr]] / 10
        condcounter[cond] <- condcounter[cond] + 1
        for (ch in 1:64){
        temp <- as.matrix(EEGdata[(triggertimes[tr]:(triggertimes[tr] + samplerate * duration - 1))-1499, ch + 2])
        temp <- temp - mean(temp[501:1500])
        allsamples[condcounter[cond],cond,ch,1:(1000*duration)] <- as.vector(rowMeans(temp))
        }
      }
    }
allsamples[which(is.na(allsamples))] <- 0     # remove any NaN values
meanwaveforms <- array(0,dim=c(14,64,14000))
trialsincluded <- matrix(0,nrow=14,ncol=64)
    for (cond in 1:14){
      for (ch in 1:64){
        temp <- allsamples[1:condcounter[cond],cond,ch,]
        allspec <- matrix(0,nrow=condcounter[cond],ncol=10000)
        for (trial in 1:condcounter[cond]){
          allspec[trial,] <- fft(temp[trial,2501:12500])/10000
        }
        targetresps <- allspec[,tindex]
        maskresps <- allspec[,mindex]
      if (sum(abs(targetresps))>0){  
      tempxy <- data.frame(Re(targetresps),Im(targetresps))
      D <- sqrt(mahalanobis(tempxy,colMeans(tempxy),cov(tempxy)))
      i1 <- (D<3)
      tempxy <- data.frame(Re(maskresps),Im(maskresps))
      D <- sqrt(mahalanobis(tempxy,colMeans(tempxy),cov(tempxy)))
      i2 <- (D<3)
      i <- which((i1*i2)>0)
        meanwaveforms[cond,ch,] <- colMeans(temp[i,])
        trialsincluded[cond,ch] <- length(i)
      }
      }
    }

@@

@@

d <- dir(path = 'temp/VilidaiteProcessed', pattern = '*.RData', full.names=TRUE)
EEGtimecourse <- array(0,dim=c(length(d),length(findex),3,13000))
EEGspectra <- array(0,dim=c(length(d),3,20*10))
EEGscalpresps <- array(0,dim=c(length(d),4,14,64))
EEGallconds <- array(0,dim=c(length(d),4,14))

for (s in 1:length(d)){
  load(d[s])
  targetchannels <- match(targetelectrodes, electrodes) - 2

@@

save(file='temp/Pilotsummary.RData',list=c('EEGtimecourse','electrodes','EEGscalpresps','EEGspectra','EEGallconds'))
}

@@

load('temp/Pilotsummary.RData')
nsubjs <- dim(EEGtimecourse)[1]
EEGmeantimes <- array(0,dim=c(length(findex),3,13000))
EEGSEtimes <- array(0,dim=c(length(findex)+1,3,13000))
EEGmaskindices <- array(0,dim=c(length(findex)+1,13000))
EEGmaskSE <- array(0,dim=c(length(findex)+1,13000))
EEGmeanscalp <- abs(apply(abs(EEGscalpresps),2:4,mean))
EEGmeanspec <- abs(apply(abs(EEGspectra),2:3,mean))
EEGmeanconds <- apply(abs(EEGallconds),2:3,mean)
EEGcondsSE <- apply(abs(EEGallconds),2:3,sd)/sqrt(nsubjs)

@@

  pdf('temp/figures/Fig1.pdf', bg="transparent", height = 6, width = 16)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

@@

targetcond <- NULL
maskcond <- NULL
subj <- NULL
targetamp <- NULL
maskamp <- NULL
i <- 0
for (s in 1:100){
  for (cond in 1:14){
    i <- i + 1
    subj[i] <- s
    targetcond[i] <- ((cond-1) %% 7) + 1
    maskcond[i] <- ceiling(cond/7)
    targetamp[i] <- abs(EEGallconds[s,1,cond])
    maskamp[i] <- abs(EEGallconds[s,2,cond])
  }
}
anovadata <- data.frame(subj,targetcond,maskcond,targetamp,maskamp)
anovadata$subj <- as.factor(anovadata$subj)
anovadata$targetcond <- as.factor(anovadata$targetcond)
anovadata$maskcond <- as.factor(anovadata$maskcond)

@@

```{r Pilotdata, fig.cap="Summary of pilot analysis of data from Vilidiate et al. (2018). Panel (a) shows contrast response functions at the target frequency (7Hz, left) and the mask frequency (5Hz, right). Insets show the distribution of activity across the scalp, with points marking the electrodes over which signals were averaged (Oz, POz, O1 and O2). Panel (b) shows Fourier spectra for the single component stimuli and their combination (plaid). Note the strong second harmonic components at 14Hz and 10Hz. Panel (c) shows timecourses of frequency-locked responses to a single stimulus (black) and the plaid stimulus (blue), compared to baseline (grey). Panel (d) shows the timecourse of suppression at each frequency (7Hz, 5Hz, 14Hz, 10Hz) and their average (black curve). Points around y = 1.8 indicate a significantly increasing ratio. Error bars in panel (a) and shaded regions in panels (c,d) indicate ±1SE, and grey rectangles indicate the timing of stimulus presentation. The larger symbols in panel (a) indicate the conditions used for subsequent analyses.", fig.align="center", echo=FALSE}

knitr::include_graphics('temp/figures/Fig1.pdf')

@@

Our initial reanalysis was promising, however the data were noisy despite the large sample size, because each participant contributed only 8 trials (88 seconds) to each condition. We therefore preregistered two new experiments (see https://osf.io/4qudc) to investigate these effects in greater detail. These had a similar overall design to the @Vilidaite2018 study, with some small changes intended to optimise the study (see Methods). The key differences were that we used shorter trials (because there were few changes in the latter part of the trials shown in Figure \@ref(fig:Pilotdata)d), and also focussed all trials into a smaller number of conditions, such that each participant contributed 48 repetitions (288 seconds of data) to each of 4 conditions.

@@

@@

if (processdata>1){    
d <- dir(path = 'temp/EEGprocessed', pattern = '*.RData', full.names=TRUE)
EEGtimecourse5Hz <- array(0,dim=c(length(d),4,8000))
EEGtimecourse7Hz <- array(0,dim=c(length(d),4,8000))
allwaves <- array(0,dim=c(length(d),4,9000))
allspec <- array(0,dim=c(length(d),4,duration*20))
allheads5 <- array(0,dim=c(length(d),4,64))
allheads7 <- array(0,dim=c(length(d),4,64))
allAQ <- NULL
allSPQ <- NULL
for (s in 1:length(d)){
  load(d[s])
  targetchannels <- match(targetelectrodes, electrodes) - 2
  for (cond in 1:4){
    temp <- rowMeans(subjwaves[cond,,targetchannels],na.rm=TRUE)
    nainds <- which(is.na(temp))   # find occasional NaN values in the waveform
    # temp[nainds] <- (temp[nainds-1] + temp[nainds+1])/2   # interpolate adjacent values
    temp[nainds] <- 0   # replace with zeros
    allwaves[s,cond,] <- temp - mean(temp)
    spec <- fft(temp[1501:7500])/6000
    allspec[s,cond,] <- spec[1+(1:(duration*20))]

@@

save(file='temp/EEGsummary.RData',list=c('allheads5','allheads7','allspec','allwaves','EEGtimecourse5Hz','EEGtimecourse7Hz','electrodes','allAQ','allSPQ'))
 }

@@

@@

# pdf("temp/figures/Fig2.pdf", bg="transparent", height = 10, width = 12)
# 
# par(mfrow=c(3,4))

@@

  pdf('temp/figures/Fig2.pdf', bg="transparent", height = 6, width = 15)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

@@

knitr::include_graphics('temp/figures/Fig2.pdf')

@@

@@

@@

  e1 <- readTIFF('temp/figures/groupbrainbothSNR.tif')
  e2 <- readTIFF('temp/figures/LH.tif')
  e3 <- readTIFF('temp/figures/RH.tif')

  pdf('temp/figures/Fig3.pdf', bg="transparent", height = 6, width = 15)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

knitr::include_graphics('temp/figures/Fig3.pdf')

@@

  pdf('temp/figures/Fig4.pdf', bg="transparent", height = 10, width = 12)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

knitr::include_graphics('temp/figures/Fig4.pdf')

@@

One important question is whether the dynamic increase in suppression can be explained by the stimulus onset transient. This is a possibility that cannot be ruled out for some of our data. For example, the steep increase in suppression in Figure \@ref(fig:EEGdata)f has a similar timecourse to the onset transient in Figure \@ref(fig:EEGdata)d. However, there are also counterexamples where suppression continues to increase well beyond the first 1 second of stimulus presentation (e.g. Figure \@ref(fig:EEGdata)e). It is currently unclear why there appear to be such substantial differences between temporal frequency conditions, especially with such similar frequencies (5 and 7Hz). However the differences are relatively consistent across experiments. For example, 5Hz flicker produces a more gradual increase in suppression across all three data sets, compared with 7Hz flicker. These differences may be a consequence of visual channels with different temporal tuning interacting with the stimulation frequency, as well as any nonlinearities that govern suppression. Or there could be an asymmetry, whereby the relative temporal frequency between the two stimulus components affects the character of suppression [@Liza2022]. We hope that in future dynamic models of early vision [e.g. @Zhou2019] will be able to explain these differences.

@@

In Experiments 1 and 2, stimuli were presented using a ViewPixx LCD display device (VPixx Technologies, Canada) with a resolution of 1920 $\times$ 1080 pixels and a refresh rate of 120Hz. The display was gamma corrected using a Minolta LS110 photometer. In Experiment 2, participants wore active stereo shutter glasses (NVidia 3D Vision 2) that were synchronised with the display using an infra-red signal. EEG data were collected using a 64-channel Waveguard cap, and were amplified and digitised at 1000Hz using an ANT Neuroscan system. Electrode impedance was maintained below 5k$\Omega$, and referenced to a whole-head average.

@@

EEG data from Experiments 1 and 2 were first imported into Matlab using components of the EEGlab toolbox [@Delorme2004], and converted into a compressed ASCII format. Primary data analysis was then conducted using a bespoke _R_ script. In brief, we epoched each trial and extracted the average timecourse across four occipital electrodes (\emph{Oz}, \emph{POz}, \emph{O1} and \emph{O2}), and then calculated the Fourier transform of this average waveform. We excluded trials for which the Mahalanobis distance of the complex Fourier components exceeded 3 [for details see @Baker2021]. This resulted in % of trials being excluded for Experiment 1, % of trials for Experiment 2, and % of trials for Experiment 3. Next we averaged the waveforms across all remaining trials, and calculated the Fourier transform in a 1-second sliding window to generate timecourses for each participant. We divided the timecourse for the target-only condition by the timecourse for the target + mask condition to produce a suppression ratio. These were then converted to logarithmic (dB) units for averaging, calculation of standard errors, and statistical comparisons. For display purposes we smoothed the timecourses using a cubic spline function, however all statistical comparisons used the unsmoothed data.

>>>>

# reasonably compact code to check which packages are installed, install the missing ones, and activate all
packagelist <- c('knitr','remotes','tictoc','R.matlab','bookdown','rmarkdown','grImport','tiff','pals','ez','gtools','pracma') # list of CRAN packages
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
if (!'osfr' %in% installed.packages()[,1]){remotes::install_github("centerforopenscience/osfr")}
if (!'FourierStats' %in% installed.packages()[,1]){remotes::install_github("bakerdh/FourierStats")}
packagelist <- c(packagelist,'osfr','FourierStats')
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

    # compare each cluster to the null distribution, retain the significant ones
    clusterps <- 0*(1:nclusts)
    if (length(nulldist)==nresamples){
    for (cc in 1:nclusts){
      i <- which(abs(nulldist)>abs(sumtvals[cc]))
      if (!pracma::isempty(i)){clusterps[cc] <- length(i)/nresamples}
      if (clusterps[cc]<clustthresh){
        sigclustercount <- sigclustercount + 1
        clusterpoints[[sigclustercount]] <- clusterlist[[cc]]
        clusterpout[sigclustercount] <- clusterps[cc]
        }
    }}
  }
clustout <- NULL
clustout$clusterpoints <- clusterpoints
clustout$nclusters <- sigclustercount
clustout$pvals <- clusterpout

@@

# check if directories exist to store data files (ignored by git)
if (!dir.exists('temp/')){dir.create('temp/')}
if (!dir.exists('temp/VilidaiteProcessed/')){dir.create('temp/VilidaiteProcessed/')}
if (!dir.exists('temp/raw/')){dir.create('temp/raw/')}
if (!dir.exists('temp/raw/temp/')){dir.create('temp/raw/temp/')}
if (!dir.exists('temp/MEGdata/')){dir.create('temp/MEGdata/')}
if (!dir.exists('temp/EEGprocessed/')){dir.create('temp/EEGprocessed/')}

@@

Normalization itself is not a single process. Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation [@Heeger1992; @Foley1994], eye-of-origin [@Legge1979; @Sengpiel1994] and spatial position [@Petrov2005; @Cannon1991]. At present there is evidence of normalization reweighting between stimuli with orthogonal orientations [@Aschner2018], and adjacent spatial positions [@Yiltiz2020]. We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways. This is plausible, given that suppression within and between the eyes has different spatiotemporal tuning [@Meese2009], and dichoptic masking can be reduced by adapting to the mask [@Sengpiel2005; @Li2005; @Baker2007], unlike within-eye masking. Dynamic fluctuations in interocular suppression are a key feature of binocular rivalry [@Wilson2003; @Alais2010], for which autism-related differences have also been reported [@Robertson2013]. By comparing monocular and dichoptic suppression, we are also able to isolate any contribution of early pre-cortical factors. This is because interocular suppression impacts in primary visual cortex, and therefore bypasses earlier retinal and subcortical stages of processing that may contribute to monocular suppression [@Li2005].

@@

Our recent work has identified a potential autism-related difference in gain control using steady-state EEG methods. @Vilidaite2018 measured contrast response functions for flickering stimuli in a large group of neurotypical adults, who were grouped according to their autism quotient [AQ, see @Baron-Cohen2001] score. The responses were comparable between high and low AQ groups at the flicker frequency of the stimulus. However at the second harmonic frequency (twice the flicker frequency), the high AQ group showed weaker responses than the low AQ group. This effect was replicated in an independent sample of 12 adults with an autism diagnosis (compared with neurotypical controls). Because second harmonic responses are caused by nonlinear interactions in the visual system (they are absent in a linear system), and divisive suppression is a nonlinearity, this potentially implicates differences in suppression in autism. Furthermore, there appears to be a developmental trajectory, as autistic children tested as part of the same study [@Vilidaite2018] showed weaker responses at both the first and second harmonic frequencies.

@@

# single participant analysis of data from Vilidaite et al. (2018)

@@

datadir <- 'temp/raw/'
tempdir <- 'temp/raw/temp/'

@@

      for (tr in 1:counter) {
        cond <- EEGdata$Trigger[triggertimes[tr]] / 10
        condcounter[cond] <- condcounter[cond] + 1
        for (ch in 1:64){
        temp <- as.matrix(EEGdata[(triggertimes[tr]:(triggertimes[tr] + samplerate * duration - 1))-1499, ch + 2])
        temp <- temp - mean(temp[501:1500])  # subtract mean of 1 second pre-trial baseline average
        allsamples[condcounter[cond],cond,ch,1:(1000*duration)] <- as.vector(rowMeans(temp))
        }
      }
    }
allsamples[which(is.na(allsamples))] <- 0     # remove any NaN values
meanwaveforms <- array(0,dim=c(14,64,14000))
trialsincluded <- matrix(0,nrow=14,ncol=64)
    for (cond in 1:14){
      for (ch in 1:64){
        temp <- allsamples[1:condcounter[cond],cond,ch,]
        allspec <- matrix(0,nrow=condcounter[cond],ncol=10000)
        for (trial in 1:condcounter[cond]){
          allspec[trial,] <- fft(temp[trial,2501:12500])/10000
        }
        targetresps <- allspec[,tindex]
        maskresps <- allspec[,mindex]
      if (sum(abs(targetresps))>0){  
      tempxy <- data.frame(Re(targetresps),Im(targetresps))
      D <- sqrt(mahalanobis(tempxy,colMeans(tempxy),cov(tempxy)))
      i1 <- (D<3)
      tempxy <- data.frame(Re(maskresps),Im(maskresps))
      D <- sqrt(mahalanobis(tempxy,colMeans(tempxy),cov(tempxy)))
      i2 <- (D<3)
      i <- which((i1*i2)>0)
        meanwaveforms[cond,ch,] <- colMeans(temp[i,])
        trialsincluded[cond,ch] <- length(i)
      }
      }
    }

@@

# group level analysis of data from Vilidaite et al. (2018)

@@

d <- dir(path = 'temp/VilidaiteProcessed', pattern = '*.RData', full.names=TRUE)
EEGtimecourse <- array(0,dim=c(length(d),length(findex),3,13000))
EEGspectra <- array(0,dim=c(length(d),3,20*10))
EEGscalpresps <- array(0,dim=c(length(d),4,14,64))
EEGallconds <- array(0,dim=c(length(d),4,14))
alltrialcounts <- array(0,dim=c(length(d),14,64))

for (s in 1:length(d)){
  load(d[s])
  targetchannels <- match(targetelectrodes, electrodes) - 2
  alltrialcounts[s,,] <- trialsincluded

@@

save(file='temp/Pilotsummary.RData',list=c('EEGtimecourse','electrodes','EEGscalpresps','EEGspectra','EEGallconds','alltrialcounts'))
}

# plot data from Vilidaite et al. (2018)

@@

nsubjs <- dim(EEGtimecourse)[1]
EEGmeantimes <- array(0,dim=c(length(findex),3,13000))
EEGSEtimes <- array(0,dim=c(length(findex)+1,3,13000))
EEGmaskindices <- array(0,dim=c(length(findex)+1,13000))
EEGmaskSE <- array(0,dim=c(length(findex)+1,13000))
EEGmeanscalp <- abs(apply(abs(EEGscalpresps),2:4,mean))
EEGmeanspec <- abs(apply(abs(EEGspectra),2:3,mean))
EEGmeanconds <- apply(abs(EEGallconds),2:3,mean)
EEGcondsSE <- apply(abs(EEGallconds),2:3,sd)/sqrt(nsubjs)

@@

  pdf('Figures/VilidaiteData.pdf', bg="transparent", height = 6, width = 16)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

# do statistics for Vilidiate et al. data

@@

Pilotpercentexcluded <- 100 - (100*mean(alltrialcounts[,condlist,28:31])/8)
targetcond <- NULL
maskcond <- NULL
subj <- NULL
targetamp <- NULL
maskamp <- NULL
i <- 0
for (s in 1:100){
  for (cond in 1:14){
    i <- i + 1
    subj[i] <- s
    targetcond[i] <- ((cond-1) %% 7) + 1
    maskcond[i] <- ceiling(cond/7)
    targetamp[i] <- abs(EEGallconds[s,1,cond])
    maskamp[i] <- abs(EEGallconds[s,2,cond])
  }
}
anovadata <- data.frame(subj,targetcond,maskcond,targetamp,maskamp)
anovadata$subj <- as.factor(anovadata$subj)
anovadata$targetcond <- as.factor(anovadata$targetcond)
anovadata$maskcond <- as.factor(anovadata$maskcond)

@@

```{r Pilotdata, fig.cap="Summary of pilot analysis of data from Vilidiate et al. (2018). Panel (a) shows contrast response functions at the target frequency (7Hz, left) and the mask frequency (5Hz, right). Insets show the distribution of activity across the scalp, with points marking the electrodes over which signals were averaged (Oz, POz, O1 and O2). Panel (b) shows Fourier spectra for the single component stimuli and their combination (plaid). Note the strong second harmonic components at 14Hz and 10Hz. Panel (c) shows timecourses of frequency-locked responses to a single stimulus (black) and the plaid stimulus (blue), compared to baseline (grey). Panel (d) shows the timecourse of suppression at each frequency (7Hz, 5Hz, 14Hz, 10Hz) and their average (black curve). Points around y = 1.8 indicate a significantly increasing ratio. Error bars in panel (a) and shaded regions in panels (c,d) indicate ±1SE across N=100 participants, and grey rectangles indicate the timing of stimulus presentation. The larger symbols in panel (a) indicate the conditions used for subsequent analyses.", fig.align="center", echo=FALSE}

knitr::include_graphics('Figures/VilidaiteData.pdf')

@@

Our initial reanalysis was promising, however the data were noisy despite the large sample size (of N=100), because each participant contributed only 8 trials (88 seconds) to each condition. We therefore preregistered two new experiments (see https://osf.io/4qudc) to investigate these effects in greater detail. These had a similar overall design to the @Vilidaite2018 study, with some small changes intended to optimise the study (see Methods). The key differences were that we used shorter trials (because there were few changes in the latter part of the trials shown in Figure \@ref(fig:Pilotdata)d), and also focussed all trials into a smaller number of conditions, such that each participant contributed 48 repetitions (288 seconds of data) to each of 4 conditions.

@@

# analyse raw data from EEG study

@@

# average EEG data across participants

d <- dir(path = 'temp/EEGprocessed', pattern = '*.RData', full.names=TRUE)
EEGtimecourse5Hz <- array(0,dim=c(length(d),4,8000))
EEGtimecourse7Hz <- array(0,dim=c(length(d),4,8000))
allwaves <- array(0,dim=c(length(d),4,9000))
allspec <- array(0,dim=c(length(d),4,duration*20))
allheads5 <- array(0,dim=c(length(d),4,64))
allheads7 <- array(0,dim=c(length(d),4,64))
EEGtrialsincluded <- array(0,dim=c(length(d),4,64))
allAQ <- NULL
allSPQ <- NULL
for (s in 1:length(d)){
  load(d[s])
  targetchannels <- match(targetelectrodes, electrodes) - 2
  EEGtrialsincluded[s,,] <- trialsincluded
  for (cond in 1:4){
    temp <- rowMeans(subjwaves[cond,,targetchannels],na.rm=TRUE)
    nainds <- which(is.na(temp))   # find occasional NaN values in the waveform
    # temp[nainds] <- (temp[nainds-1] + temp[nainds+1])/2   # interpolate adjacent values
    temp[nainds] <- 0   # replace with zeros
    allwaves[s,cond,] <- temp - mean(temp)
    spec <- fft(temp[1501:7500])/6000
    allspec[s,cond,] <- spec[1+(1:(duration*20))]

@@

save(file='temp/EEGsummary.RData',list=c('allheads5','allheads7','allspec','allwaves','EEGtimecourse5Hz','EEGtimecourse7Hz','electrodes','allAQ','allSPQ','EEGtrialsincluded'))
 }

# bootstrapping and figure plotting for EEG data

@@

EEGpercentexcluded <- 100 - mean(100*EEGtrialsincluded[,,28:31]/48)

@@

@@

  pdf('Figures/EEGdata.pdf', bg="transparent", height = 6, width = 15)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

# statistical testing

@@

knitr::include_graphics('Figures/EEGdata.pdf')

@@

# combine MEG data across participants

@@

# plot MEG figure

@@

  e1 <- readTIFF('temp/groupbrainbothSNR.tif')
  e2 <- readTIFF('temp/LH.tif')
  e3 <- readTIFF('temp/RH.tif')

  pdf('Figures/MEGdata.pdf', bg="transparent", height = 6, width = 15)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

knitr::include_graphics('Figures/MEGdata.pdf')

@@

  pdf('Figures/AQfigure.pdf', bg="transparent", height = 10, width = 12)
  par(mar=c(0.1,0.1,0.1,0.1))
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

knitr::include_graphics('Figures/AQfigure.pdf')

@@

One important question is whether the dynamic increase in suppression can be explained by the stimulus onset transient. This is a possibility that cannot be ruled out for some of our data. For example, the steep increase in suppression in Figure \@ref(fig:EEGdata)f has a similar timecourse to the onset transient in Figure \@ref(fig:EEGdata)d. However, there are also counterexamples where suppression continues to increase well beyond the first 1 second of stimulus presentation (e.g. Figure \@ref(fig:EEGdata)e). It is currently unclear why there appear to be such substantial differences between temporal frequency conditions, especially with such similar frequencies (5 and 7Hz). However the differences are relatively consistent across experiments. For example, 5Hz flicker produces a more gradual increase in suppression across all three data sets, compared with 7Hz flicker. These differences may be a consequence of visual channels with different temporal tuning interacting with the stimulation frequency, as well as any nonlinearities that govern suppression. Or there could be an asymmetry, whereby the relative temporal frequency between the two stimulus components affects the character of suppression [@Liza2022]. We hope to be able to model these effects in the future, for example by using dynamic models of early vision that incorporate time-lagged gain control [e.g. @Zhou2019].

We did not observe clear differences in the timecourse between monocular and dichoptic suppression. This is important, because the dichoptic arrangement bypasses early stages of processing before the cortex (e.g. the retina and lateral geniculate nucleus). It suggests that the dynamic increases in suppression occur in the cortex, consistent with our MEG data that find evidence of reweighting in V1 (see Fig \@ref(fig:MEGdata)), and with previous neurophysiological work [@Aschner2018]. It is currently unclear whether these effects originate in V1, or might involve feedback from higher areas. The similarity between monocular and dichoptic effects also differs from work on adaptation to individual mask components. In both physiological [@Li2005; @Sengpiel2005] and psychophysical [@Baker2007] paradigms, adapting to a dichoptic mask reduces its potency, whereas adapting to a monocular mask has little or no effect. Normalization reweighting offers an explanation for why monocular masks presented in isolation do not adapt: if suppressive weights are determined by co-occurrence of stimuli, presentation of an isolated mask will have little effect. However this cannot explain the dichoptic adaptation effects without invoking additional binocular processes, such as competition between summing and differencing channels [e.g. @May2012].

@@

In Experiments 1 and 2, stimuli were presented using a ViewPixx 3D LCD display device (VPixx Technologies, Canada) with a resolution of 1920 $\times$ 1080 pixels and a refresh rate of 120Hz. The display was gamma corrected using a Minolta LS110 photometer. In Experiment 2, participants wore active stereo shutter glasses (NVidia 3D Vision 2) that were synchronised with the display using an infra-red signal. EEG data were collected using a 64-channel Waveguard cap, and were amplified and digitised at 1000Hz using an ANT Neuroscan system. Electrode impedance was maintained below 5k$\Omega$, and referenced to a whole-head average.

@@

EEG data from Experiments 1 and 2 were first imported into Matlab using components of the EEGlab toolbox [@Delorme2004], and converted into a compressed ASCII format. Primary data analysis was then conducted using a bespoke _R_ script. In brief, we epoched each trial and extracted the average timecourse across four occipital electrodes (\emph{Oz}, \emph{POz}, \emph{O1} and \emph{O2}), and then calculated the Fourier transform of this average waveform. We excluded trials for which the Mahalanobis distance of the complex Fourier components exceeded 3 [for details see @Baker2021]. This resulted in `r round(Pilotpercentexcluded,digits=2)`% of trials being excluded for Experiment 1, `r round(EEGpercentexcluded,digits=2)`% of trials for Experiment 2, and `r round(EEGpercentexcluded,digits=2)`% of trials for Experiment 3. Next we averaged the waveforms across all remaining trials, and calculated the Fourier transform in a 1-second sliding window to generate timecourses for each participant. We divided the timecourse for the target-only condition by the timecourse for the target + mask condition to produce a suppression ratio. These were then converted to logarithmic (dB) units for averaging, calculation of standard errors, and statistical comparisons. For display purposes we smoothed the timecourses using a cubic spline function, however all statistical comparisons used the unsmoothed data.
